---
title: "Homework 2"
subtitle: "Submission 1, Spring 2026"
author: "Sophie Kim"
format:
  pdf:
    output-file: "Kim-S-hwk2-1"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---

```{r}
#| include: false
library(tidyverse)
library(stringr)
library(readr)
library(fs)
```

```{r}
#| include: false
data.2014 <- read_csv("..//data/output/data-2014.csv")
data.2015 <- read_csv("..//data/output/data-2015.csv")
data.2016 <- read_csv("..//data/output/data-2016.csv")
data.2017 <- read_csv("..//data/output/data-2017.csv")
data.2018 <- read_csv("..//data/output/data-2018.csv")
```

```{r}
#| include: false
combined <- bind_rows(
  data.2014,
  data.2015,
  data.2016,
  data.2017,
  data.2018
)
```

```{r}
#| include: false
glimpse(combined)
```

Problem 1

```{r}
#| include: false
plan_counts <- combined %>% 
count(county, year.x, name = "plan_count")
```

```{r}
#| echo: false
boxplot(plan_count ~ year.x, data = plan_counts,
        xlab = "Year",
        ylab = "Plan count",
        main = "Distribution of plan counts by year" )
```

```{r}
#| include: false
combined <- combined %>% mutate (basic_premium = case_when(
        rebate_partc > 0 ~ 0,
        partd == "No" & !is.na(premium) & is.na(premium_partc) ~ premium,
        TRUE ~ premium_partc
      ),
      bid = case_when(
        rebate_partc == 0 & basic_premium > 0 ~ (payment_partc + basic_premium) / riskscore_partc,
        rebate_partc > 0  | basic_premium == 0 ~  payment_partc / riskscore_partc,
        TRUE ~ NA_real_
      )
                                 )
```

```{r}
#| echo: false
ggplot(filter(combined, year.x == 2014, !is.na(bid)), aes(x = bid)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Medicare Advantage Plan Bids (2014)",
    x = "Plan bid",
    y = "Frequency"
  ) 
```

```{r}
#| echo: false
ggplot(filter(combined, year.x == 2018, !is.na(bid)), aes(x = bid)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Medicare Advantage Plan Bids (2018)",
    x = "Plan bid",
    y = "Frequency"
  ) 
```

Problem 3

```{r}
#| include: false
combined_hhi <- combined %>%
  filter(!is.na(avg_enrolled), avg_enrolled > 0) %>%
  filter(!is.na(avg_enrollment), avg_enrollment > 0) %>%
  group_by(fips, year.x) %>%
  mutate(market_share = avg_enrollment / avg_enrolled
  ) %>%
  ungroup()
```

```{r}
#| include: false
county_hhi <- combined_hhi %>%
  group_by(fips, year.x) %>%
  summarise(
    hhi = sum(market_share^2),
    .groups = "drop"
  )
```

```{r}
#| include: false
avg_hhi <- county_hhi %>%
  group_by(year.x) %>%
  summarise(
    avg_hhi = mean(hhi),
    .groups = "drop"
  )
```

```{r}
avg_hhi
```

```{r}
#| echo: false
ggplot(avg_hhi, aes(x = year.x, y = avg_hhi)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Medicare Advantage Market Concentration (HHI)",
    x = "Year",
    y = "Average HHI"
  )
```

Problem 4

```{r}
#| include: false
ma_year <- combined %>%
  group_by(year.x) %>%
    mutate(ma_share = avg_enrolled / avg_eligibles) %>%
    summarise(mean = mean(ma_share, na.rm=TRUE), groups = "drop")
```

```{r}
ma_year
```

```{r}
#| echo: false
ggplot(ma_year, aes(x = year.x, y = mean)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Share of Medicare Advantage over Time",
    x = "Year",
    y = "Average Share"
  )
```

```{r}
#| include: false
glimpse(data.2018)
```

Problem 5

```{r}
#| include: false
marketshare2018 <- data.2018 %>%
  filter(!is.na(avg_enrolled), avg_enrolled > 0) %>%
  filter(!is.na(avg_enrollment), avg_enrollment > 0) %>%
  group_by(fips, year.x) %>%
  mutate(market_share = avg_enrollment / avg_enrolled
  ) %>%
  ungroup()
```

```{r}
#| include: false
hhi_2018 <- marketshare2018 %>%
  group_by(fips) %>%
  mutate(
    hhi = sum(market_share^2, na.rm = TRUE),
  )%>%
  ungroup()
```

```{r}
#| include: false
glimpse(hhi_2018)
```

```{r}
#| include: false
hhi_33 <- quantile(hhi_2018$hhi, 0.33, na.rm = TRUE)
hhi_66 <- quantile(hhi_2018$hhi, 0.66, na.rm = TRUE)
```

```{r}
#| include: false
hhi_2018 <- hhi_2018 %>%
  mutate(
    market_type = case_when(
      hhi <= hhi_33 ~ "Competitive",
      hhi >= hhi_66 ~ "Uncompetitive",
      TRUE ~ NA_character_
    )
  )
```

```{r}
#| include: false
avg_bid_by_market <- hhi_2018 %>%
  filter(!is.na(market_type), !is.na(bid)) %>%
  group_by(market_type) %>%
  summarise(
    avg_bid = mean(bid),
    .groups = "drop"
  )
```

```{r}
print(avg_bid_by_market)
```

Problem 6

```{r}
#| include: false
ffs.2018 <- read_csv("..//data/output/data-2018-ffs.csv")
```

```{r}
#| include: false
colnames(ffs.2018)
```

```{r}
#| include: false
ffs.2018 <- ffs.2018 %>%
  mutate(
    ffscost_q = ntile(avg_ffscost, 4)
  )
```

```{r}
#| include: false
q1 <- ffs.2018 %>%
  mutate(treated = if_else(ffscost_q == 1, "Treated", "Control")) %>%
  group_by(treated) %>%
  summarise(avg_bid = mean(bid, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(quartile = "Q1")

q2 <- ffs.2018 %>%
  mutate(treated = if_else(ffscost_q == 2, "Treated", "Control")) %>%
  group_by(treated) %>%
  summarise(avg_bid = mean(bid, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(quartile = "Q2")

q3 <- ffs.2018 %>%
  mutate(treated = if_else(ffscost_q == 3, "Treated", "Control")) %>%
  group_by(treated) %>%
  summarise(avg_bid = mean(bid, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(quartile = "Q3")

q4 <- ffs.2018 %>%
  mutate(treated = if_else(ffscost_q == 4, "Treated", "Control")) %>%
  group_by(treated) %>%
  summarise(avg_bid = mean(bid, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(quartile = "Q4")
```

```{r}
big_avg_table <- bind_rows(q1, q2, q3, q4) %>%
  select(quartile, treated, avg_bid, n)

big_avg_table
```

Problem 7

```{r}
#| eval: false
#| echo: true
options(repos = c(CRAN = "https://cran.rstudio.com/"))
install.packages("MatchIt")
```

```{r}
#| eval: false
#| echo: true
ate_results <- lapply(1:4, function(q) {

dat_q <- ffs.2018 %>%
    mutate(
      treat = ifelse(ffscost_q == q, 1, 0),
      outcome = bid,
      covar = avg_ffscost
    )

dat_q <- dat_q %>% 
  filter(!is.na(outcome), !is.na(treat), !is.na(covar))

nn2 <- Matching::Match(
    Y = dat_q$outcome,
    Tr = dat_q$treat,
    X = dat_q$covar,
    M = 1,
    Weight = 1,
    estimand = "ATE",
    ties=FALSE
  )

  data.frame(
    quartile = q,
    ATE = nn2$est
  )
})

ate_table <- do.call(rbind, ate_results)

print(ate_table)
```

```{r}
#| eval: false
#| echo: true
print(ate_table)
```

```{r}
#| eval: false
#| echo: true
ate_results <- lapply(1:4, function(q) {

dat_q <- ffs.2018 %>%
    mutate(
      treat = ifelse(ffscost_q == q, 1, 0),
      outcome = bid,
      covar = avg_ffscost
    )

dat_q <- dat_q %>% 
  filter(!is.na(outcome), !is.na(treat), !is.na(covar))

nn2 <- Matching::Match(
    Y = dat_q$outcome,
    Tr = dat_q$treat,
    X = dat_q$covar,
    M = 1,
    Weight = 2,
    estimand = "ATE",
    ties=FALSE
  )

  data.frame(
    quartile = q,
    ATE = nn2$est
  )
})

ate_table_2 <- do.call(rbind, ate_results)

print(ate_table_2)
```

```{r}
#| eval: false
#| echo: true
print(ate_table_2)
```

```{r}
#| eval: false
#| echo: true
ate_results <- lapply(1:4, function(q) {

  dat_q <- ffs.2018 %>%
    mutate(
      w = ifelse(ffscost_q == q, 1, 0),   
      y = bid,
      x = ffscost_q                      
    ) %>%
    filter(!is.na(y), !is.na(w), !is.na(x))

  logit.model <- glm(w ~ factor(x), family = binomial, data = dat_q)
  ps <- fitted(logit.model)

  dat_q <- dat_q %>%
    mutate(ipw = case_when(
      w == 1 ~ 1 / ps,
      w == 0 ~ 1 / (1 - ps),
      TRUE ~ NA_real_
    ))

  mean.w1 <- dat_q %>%
    filter(w == 1) %>%
    summarise(mean_y = weighted.mean(y, ipw))

  mean.w0 <- dat_q %>%
    filter(w == 0) %>%
    summarise(mean_y = weighted.mean(y, ipw))

  data.frame(
    quartile = q,
    ATE_IPW = mean.w1$mean_y - mean.w0$mean_y
  )
})

ate_table_ipw <- do.call(rbind, ate_results)
print(ate_table_ipw)
```

```{r}
#| eval: false
#| echo: true
print(ate_table_ipw)
```

```{r}
#| eval: false
#| echo: true
ate_results_reg <- lapply(1:4, function(q) {

  select.dat <- ffs.2018 %>%
    mutate(
      w = ifelse(ffscost_q == q, 1, 0),
      y = bid,
      x = ffscost_q
    ) %>%
    filter(!is.na(y), !is.na(w), !is.na(x))

  reg <- lm(y ~ w * factor(x), data = select.dat)

  pred1 <- predict(reg, newdata = transform(select.dat, w = 1))
  pred0 <- predict(reg, newdata = transform(select.dat, w = 0))

  ate <- mean(pred1 - pred0, na.rm = TRUE)

  data.frame(
    quartile = q,
    ATE_regression = ate
  )
})

ate_table_reg <- do.call(rbind, ate_results_reg)
print(ate_table_reg)
```

```{r}
#| eval: false
#| echo: true
print(ate_table_reg)
```

```{r}
#| eval: false
#| echo: true
saveRDS(ate_table_reg, "ate_table_reg.rds")
```

Problem 8

Nearest neighbor matching with inverse variance distance and with Mahalanobis distance have very similar average treatment effect values. However, when using inverse propensity weighting and simple linear regression, both methods output noticeably different values. 

Problem 9

```{r}
#| eval: false
#| echo: true
ate_results_reg_cont <- lapply(1:4, function(q) {

  select.dat <- ffs.2018 %>%
    mutate(
      w = ifelse(ffscost_q == q, 1, 0),
      y = bid,
      x1 = avg_ffscost,
      x2 = n_enrol
    ) %>%
    filter(!is.na(y), !is.na(w), !is.na(x1), !is.na(x2))

  reg <- lm(y ~ w * (x1 + x2), data = select.dat)

  pred1 <- predict(reg, newdata = transform(select.dat, w = 1))
  pred0 <- predict(reg, newdata = transform(select.dat, w = 0))

  ate <- mean(pred1 - pred0, na.rm = TRUE)

  data.frame(
    quartile = q,
    ATE_regression_continuous = ate
  )
})

ate_table_reg_cont <- do.call(rbind, ate_results_reg_cont)
print(ate_table_reg_cont)
```

```{r}
#| eval: false
#| echo: true
print(ate_table_reg_cont)
```


Problem 10

Working with this data allowed me to see how different estimators will give you vastly different values. One thing that I learned was that it is crucial to know what each name of the row represents so that there is no issue missing a part of the dataset. One thing that surprised me was how easy it was for library or model choices to cause errors or slow runtimes.


