---
title: "Homework 2"
subtitle: "Submission 2, Spring 2026"
author: "Sophie Kim"
format:
  pdf:
    output-file: "Kim-S-hwk2-2"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---


```{r}
#| include: false
library(tidyverse)
library(stringr)
library(readr)
library(fs)
```

```{r}
#| include: false
data.2014 <- read_csv("..//data/output/data-2014-new.csv")
data.2015 <- read_csv("..//data/output/data-2015.csv")
data.2016 <- read_csv("..//data/output/data-2016.csv")
data.2017 <- read_csv("..//data/output/data-2017.csv")
data.2018 <- read_csv("..//data/output/data-2018-ffs.csv")
data.2019 <- read_csv("..//data/output/data-2019.csv")
```

```{r}
#| include: false
combined <- bind_rows(
  data.2014,
  data.2015,
  data.2016,
  data.2017,
  data.2018,
  data.2019
)
```

```{r}
#| include: false
names(combined)
```

```{r}
#| include: false
combined <- combined %>% mutate (basic_premium = case_when(
        rebate_partc > 0 ~ 0,
        partd == "No" & !is.na(premium) & is.na(premium_partc) ~ premium,
        TRUE ~ premium_partc
      ),
      bid = case_when(
        rebate_partc == 0 & basic_premium > 0 ~ (payment_partc + basic_premium) / riskscore_partc,
        rebate_partc > 0  | basic_premium == 0 ~  payment_partc / riskscore_partc,
        TRUE ~ NA_real_
      )
                                 )
```

```{r}
#| include: false
names(combined)
```

Problem 1

```{r}
#| include: false
combine <- combined %>%
    distinct(fips, year, contractid)
    
  # Part D but no Part C premium
  #filter(!(partd == "Yes" & is.na(premium_partc))) 
```

```{r}
#| include: false
names(combined)
```

```{r}
#| include: false
plan_counts <- combine %>%
  group_by(fips, year) %>%
  select(fips,year) %>%
  summarise(
    plan_count = n(),
      .groups = "drop"
  )

summary(plan_counts$plan_count)
max(plan_counts$plan_count)
mean(plan_counts$plan_count)
```

```{r}
#| echo: false
ggplot(plan_counts, aes(x = factor(year), y = plan_count)) +
  geom_boxplot() +
  labs(
    x = "Year",
    y = "Number of Plans per County",
    title = "Distribution of Medicare Advantage Plan Counts by County (2014–2019)"
  ) +
  theme_minimal()
```

```{r}
#| include: false
#plan_counts <- combined %>% 
#group_by(fips)%>% 
#count(fips, year, name = "plan_count")
```

```{r}
#| echo: false
#boxplot(plan_count ~ year, data = plan_counts,
        #xlab = "Year",
        #ylab = "Plan count",
        #main = "Distribution of plan counts by year" )
```

Problem 2

```{r}
#| echo: false
ggplot(filter(combined, year == 2014, !is.na(bid)), aes(x = bid)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Medicare Advantage Plan Bids (2014)",
    x = "Plan bid",
    y = "Frequency"
  ) 
```

```{r}
#| echo: false
ggplot(filter(combined, year == 2018, !is.na(bid)), aes(x = bid)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Medicare Advantage Plan Bids (2018)",
    x = "Plan bid",
    y = "Frequency"
  ) 
```

Problem 3

```{r}
#| include: false
combined_hhi <- combined %>%
  filter(!is.na(avg_enrolled), avg_enrolled > 0) %>%
  filter(!is.na(avg_enrollment), avg_enrollment > 0) %>%
  group_by(fips, year) %>%
  mutate(market_share = avg_enrollment / avg_enrolled
  ) %>%
  ungroup()
```

```{r}
#| include: false
county_hhi <- combined_hhi %>%
  group_by(fips, year) %>%
  summarise(
    hhi = sum(market_share^2),
    .groups = "drop"
  )
```

```{r}
#| include: false
avg_hhi <- county_hhi %>%
  group_by(year) %>%
  summarise(
    avg_hhi = mean(hhi),
    .groups = "drop"
  )
```

```{r}
avg_hhi
```

```{r}
#| echo: false
ggplot(avg_hhi, aes(x = year, y = avg_hhi)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Medicare Advantage Market Concentration (HHI)",
    x = "Year",
    y = "Average HHI"
  )
```

Problem 4

```{r}
#| include: false
ma_year <- combined %>%
  group_by(year) %>%
    mutate(ma_share = avg_enrolled / avg_eligibles) %>%
    summarise(
    mean_ma_share = mean(ma_share, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| echo: false
knitr::kable(
  ma_year,
  digits = 6,
  col.names = c("Year", "Mean MA Enrollment Share"),
  caption = "Average Share of Medicare Advantage (2014–2019)"
)
```

```{r}
#| echo: false
ggplot(ma_year, aes(x = year, y =  mean_ma_share)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Share of Medicare Advantage over Time",
    x = "Year",
    y = "Average Share"
  )
```

Problem 5

```{r}
#| include: false
marketshare2018 <- data.2018 %>%
  filter(!is.na(avg_enrolled), avg_enrolled > 0) %>%
  filter(!is.na(avg_enrollment), avg_enrollment > 0) %>%
  group_by(fips, year) %>%
  mutate(market_share = avg_enrollment / avg_enrolled
  ) %>%
  ungroup()
```

```{r}
#| include: false
hhi_by_county <- marketshare2018 %>%
  group_by(fips) %>%
  summarize(hhi = sum(market_share^2, na.rm = TRUE), .groups = "drop")
  # summarise(mean_bid = mean(bid, na.rm = TRUE), .groups = "drop")
```

```{r}
#| include: false
hhi_33 <- quantile(hhi_by_county$hhi, 0.33, na.rm = TRUE)
hhi_66 <- quantile(hhi_by_county$hhi, 0.66, na.rm = TRUE)
```

```{r}
#| include: false
hhi_by_county <- hhi_by_county %>%
  mutate(market_type = case_when(
    hhi <= hhi_33 ~ "Competitive",
    hhi >= hhi_66 ~ "Uncompetitive",
    TRUE ~ NA_character_
  ))
```

```{r}
#| include: false
plan_level <- marketshare2018 %>%
  left_join(hhi_by_county, by = "fips") %>%
  filter(!is.na(market_type), !is.na(bid))

county_mean_bid <- plan_level %>%
  group_by(fips, market_type) %>%
  summarise(mean_bid = mean(bid, na.rm = TRUE), .groups = "drop")
        # i want to collapse all plans in one county to one (instead of mutate)

avg_bid_by_market <- county_mean_bid %>%
  group_by(market_type) %>%
  summarise(avg_bid = mean(mean_bid), .groups = "drop")
```

```{r}
#| echo: false
knitr::kable(
  avg_bid_by_market,
  digits = 2,
  col.names = c("Market Type", "Average Bid"),
  caption = "Average Medicare Advantage Bid by Market Competitiveness (2018)"
)
```

```{r}
#| include: false
# marketshare2018 <- data.2018 %>%
#   filter(!is.na(avg_enrolled), avg_enrolled > 0) %>%
#   filter(!is.na(avg_enrollment), avg_enrollment > 0) %>%
#   group_by(fips, year) %>%
#   mutate(market_share = avg_enrollment / avg_enrolled
#   ) %>%
#   ungroup()
```

```{r}
#| include: false
# hhi_2018 <- marketshare2018 %>%
#   group_by(fips) %>%
#   mutate(
#     hhi = sum(market_share^2, na.rm = TRUE),
#   )%>%
#   ungroup()
```

```{r}
#| include: false
# avg_bid_by_market <- hhi_2018 %>%
#   filter(!is.na(market_type), !is.na(bid)) %>%
#   group_by(market_type) %>%
#   summarise(
#     avg_bid = mean(bid),
#     .groups = "drop"
#   )
```

```{r}
# print(avg_bid_by_market)
```

Problem 6

```{r}
#| include: false
# ffs.2018 <- read_csv("..//data/output/data-2018-ffs.csv")
```

```{r}
#| include: false
# colnames(ffs.2018)
```

```{r}
#| include: false
data.2018.ffs <- data.2018 %>%
  mutate(
    ffscost_q = ntile(avg_ffscost, 4)
  )
# made 4 quartiles here 
```

```{r}
#| include: false
q1 <- data.2018.ffs%>%
  mutate(treated = if_else(ffscost_q == 1, "Treated", "Control")) %>%
  group_by(treated) %>%
  summarise(avg_bid = mean(bid, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(quartile = "Q1")

q2 <-  data.2018.ffs %>%
  mutate(treated = if_else(ffscost_q == 2, "Treated", "Control")) %>%
  group_by(treated) %>%
  summarise(avg_bid = mean(bid, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(quartile = "Q2")

q3 <- data.2018.ffs %>%
  mutate(treated = if_else(ffscost_q == 3, "Treated", "Control")) %>%
  group_by(treated) %>%
  summarise(avg_bid = mean(bid, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(quartile = "Q3")

q4 <- data.2018.ffs %>%
  mutate(treated = if_else(ffscost_q == 4, "Treated", "Control")) %>%
  group_by(treated) %>%
  summarise(avg_bid = mean(bid, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(quartile = "Q4")
```

```{r}
#| echo: false
big_avg_table <- bind_rows(q1, q2, q3, q4) %>%
  select(quartile, treated, avg_bid, n)%>%
  filter(!is.na(treated))
```

```{r}
#| echo: false
knitr::kable(
  big_avg_table,
  digits = 5,
  col.names = c("Quartile", "Treated/Control", "Average Bid", "n"),
  caption = "Average Bid for Treated/Control Groups per Quartile(2018)"
)
```

Problem 7

```{r}
#| include: false
options(repos = c(CRAN = "https://cran.rstudio.com/"))
install.packages("MatchIt")
```

```{r}
#| include: false
library(MatchIt)

pooled_data <- lapply(1:4, function(q) {

  dat_q <- data.2018.ffs%>%
    mutate(
      treat = ifelse(ffscost_q == q, 1, 0),
      outcome = bid,
      covar = avg_ffscost
    ) %>%
    filter(!is.na(outcome), !is.na(treat), !is.na(covar))

  dat_q
})

pooled_data <- do.call(rbind, pooled_data)
```

```{r}
#| include: false
m.out <- matchit(
  treat ~ covar,
  data = pooled_data,
  method = "nearest",
  distance = pooled_data$covar,
  replace = FALSE
)

m.data <- match.data(m.out)

inv_ate <- mean(m.data$outcome[m.data$treat == 1]) - mean(m.data$outcome[m.data$treat== 0])
```

```{r}
#| include: false
print(inv_ate)
```

```{r}
#| include: false
m.out <- matchit(
  treat ~ covar,
  data = pooled_data,
  method = "nearest",
  distance = "mahalanobis",
  replace = FALSE
)

m.data <- match.data(m.out)
mah_ate <- mean(m.data$outcome[m.data$treat == 1]) - mean(m.data$outcome[m.data$treat== 0])
```

```{r}
#| include: false
print(mah_ate)
```

```{r}
#| include: false
logit.model <- glm(treat ~ covar, family = binomial, data = pooled_data)
pooled_data$ps <- fitted(logit.model)

pooled_data <- pooled_data %>%
  mutate(ipw = case_when(
    treat == 1 ~ 1/ps,
    treat == 0 ~ 1/(1 - ps)
  ))

mean.w1 <- pooled_data %>%
  filter(treat == 1) %>%
  summarize(mean_y = weighted.mean(outcome, ipw))

mean.w0 <- pooled_data %>%
  filter(treat == 0) %>%
  summarize(mean_y = weighted.mean(outcome, ipw))

ipw_ate <- mean.w1$mean_y - mean.w0$mean_y
```

```{r}
#| include: false
print(ipw_ate)
```

```{r}
#| include: false
reg1 <- lm(outcome ~ covar, data = pooled_data %>% filter(treat == 1))
reg0 <- lm(outcome ~ covar, data = pooled_data %>% filter(treat == 0))

pred1 <- predict(reg1, newdata = pooled_data)
pred0 <- predict(reg0, newdata = pooled_data)

reg_ate <- mean(pred1 - pred0)
     
```

```{r}
#| include: false
print(reg_ate)
```

```{r}
results <- data.frame(
  inv_ate = inv_ate,
  mah_ate = mah_ate,
  ipw_ate = ipw_ate,
  reg_ate = reg_ate
)

results
```

```{r}
#| include: false
# analysis <- data.2018.ffs %>%
#   left_join(hhi_by_county %>% select(fips, market_type), by = "fips") %>%
#   mutate(
#     high_hhi  = as.integer(market_type == "Uncompetitive"),
#     ffscost_q = as.integer(ffscost_q)
#   ) %>%
#   filter(!is.na(bid), !is.na(high_hhi), !is.na(ffscost_q))
```

```{r}
#| include: false
# get_nn_inv_q <- function(q){
#   sub <- analysis %>%
#     filter(ffscost_q == q) %>%
#     transmute(y = bid, w = high_hhi, x = bid)

#   if (length(unique(sub$w)) < 2) return(NA_real_)

#   m <- Matching::Match(
#     Y = sub$y, Tr = sub$w, X = sub$x,
#     M = 1, Weight = 1, estimand = "ATE",
#     ties = FALSE, version = "fast"
#   )

#   as.numeric(m$est)
# }

# ests <- sapply(1:4, get_nn_inv_q)
# Ns   <- sapply(1:4, function(q) sum(analysis$ffscost_q == q))

# nn_inv_ate <- weighted.mean(ests, w = Ns, na.rm = TRUE)
# nn_inv_ate
```

```{r}
#| include: false
# nrow(analysis)
# count(analysis, fips) %>% summarise(max_n = max(n))
# hhi_by_county %>%
#   count(fips) %>%
#   summarise(max_dupes = max(n))
```

```{r}
#| include: false
# market_type_by_fips <- hhi_by_county %>%
#   select(fips, market_type) %>%
#   distinct(fips, .keep_all = TRUE)

# analysis <- data.2018.ffs %>%
#   left_join(market_type_by_fips, by = "fips") %>%
#   mutate(
#     high_hhi  = if_else(market_type == "Uncompetitive", 1L, 0L),
#     ffscost_q = as.integer(ffscost_q)
#   ) %>%
#   filter(!is.na(bid), !is.na(high_hhi), !is.na(ffscost_q))
```

```{r}
#| include: false
# nrow(data.2018.ffs)
# nrow(analysis)
```

```{r}
#| include: false
# Xq <- model.matrix(~ factor(ffscost_q), data = analysis)[, -1, drop = FALSE]
```

```{r}
#| include: false
# nn_inv <- Matching::Match(
#   Y = analysis$bid,
#   Tr = analysis$high_hhi,
#   X = Xq,
#   M = 1,
#   Weight = 1,
#   estimand = "ATE",
#   exact = TRUE,
#   ties = FALSE,
#   version = "fast"
# )
```

```{r}
#| include: false
# nn_inv$est
# nn_inv$se
```

```{r}
#| include: false
# nn_mah <- Matching::Match(
#   Y = analysis$bid,
#   Tr = analysis$high_hhi,
#   X = Xq,
#   M = 1,
#   Weight = 2,
#   estimand = "ATE",
#   exact = TRUE,
#   ties = FALSE,
#   version = "fast"
# )
```

```{r}
#| include: false
# nn_mah$est
# nn_mah$se
```

```{r}
#| include: false
# analysis2 <- analysis %>% 
#   filter(!is.na(bid), !is.na(high_hhi), !is.na(ffscost_q), !is.na(avg_ffscost)) %>%
#   arrange(ffscost_q, high_hhi, fips, bid)

# nn_inv <- Matching::Match(
#   Y = analysis2$bid,
#   Tr = analysis2$high_hhi,
#   X = analysis2$avg_ffscost,   # distance variable
#   exact = analysis2$ffscost_q, # match only within quartile
#   M = 1,
#   Weight = 1,                 # inverse-variance distance
#   estimand = "ATE",
#   ties = TRUE,
#   version = "fast"
# )
```

```{r}
#| include: false
saveRDS(ate_table_reg, "ate_table_reg.rds")
```

Problem 8

Nearest neighbor matching with inverse variance distance and with Mahalanobis distance have identical average treatment effect values. However, when using inverse propensity weighting and simple linear regression, both methods output different values that are smaller. 

Problem 9

```{r}
#| include: false
pooled_data_9 <- lapply(1:4, function(q) {

  dat_q9 <- data.2018.ffs %>%
    mutate(
      treat = ifelse(ffscost_q == q, 1, 0),
      outcome = bid,
      covar1 = avg_ffscost,
      covar2 = n_enrol
    ) %>%
    filter(!is.na(outcome), !is.na(treat), !is.na(covar1), !is.na(covar2))

  dat_q9
})

pooled_data_9 <- do.call(rbind, pooled_data_9)
```

```{r}
m.mah <- matchit(
  treat ~ covar1 + covar2,
  data = pooled_data_9,
  method = "nearest",
  distance = "mahalanobis",
  replace = FALSE
)

m.man <- match.data(m.out)

mah_9<- mean(m.data$outcome[m.data$treat == 1]) - mean(m.data$outcome[m.data$treat== 0])
```

```{r}
print(mah_9)
```


Problem 10

Working with this data allowed me to see how different estimators will give you vastly different values. One thing that I learned was that it is crucial to know what each name of the row represents so that there is no issue missing a part of the dataset. One thing that surprised me was how easy it was for library or model choices to cause errors or slow runtimes.


